Transport map accelerated-PAIS, and application to inverse problems arising from multiscale stochastic reaction networks

### 2019-11-01 Resubmission/response.tex

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{amsmath,amssymb}
\usepackage{enumerate}
\usepackage{color}
\usepackage{amsthm,amsmath}
\usepackage{bbm}
\usepackage{mhchem}

%for stacking below a sum or max...
\usepackage{mathtools}

\usepackage[colorlinks]{hyperref}

\newcommand{\comment}[2]{\vspace{0.6cm}{\bf Comment:} {\it #1.}

\vspace{0.3cm}{\bf Answer:} #2}

\begin{document}
\title{Response to Referee Comments: Transport map-accelerated adaptive importance sampling}
\maketitle

Firstly, we would like to thank the referees, associate editor and editor for their careful consideration of our manuscript. We largely agree with the comments made, and have adjusted the paper to reflect the extremely helpful suggestions that have been given. In particular, both referees commented on the long length of the paper and the somewhat disjoint feel between the two main topics discussed within it. In order to address this primary point, we have made substantial reductions to the paper in the areas suggested by the referees. We have also reduced the emphasis on the multiscale stochastic reaction network material, aiming to use this as an example of a nice application for the methods outlined, as opposed to an equally-weighted component of the work. We hope that this has gone a long way to knitting the two parts of the work together in a way which tells a more coherent story.

Below we list the comments made by the referees and respond to them in turn, highlighting where we have made changes in the manuscript to address them. We also attach a version of the paper with the changes in text highlighted in red. Since a lot of the larger changes involved removal of sections of text and material, we have also attached a copy of the originally submitted paper with the removed sections highlighted in blue.

We would like to thank the referees and editors for their time in considering our manuscript for publication in SIAM JUQ.

\section*{Referee 1 comments}

\comment{Length and scope. Not including references, this manuscript is 32 pages long and seems to have two somewhat disjoint parts. The first part develops the transport map accelerated importance sampler and provides details of the algorithm. The second part seems to focus more on multiscale approximations of stochastic reaction networks, and the impact that these approximations have on the posterior distribution. While interesting, especially within a model discrepancy context, I feel that this part of the paper is tangential to the paper's main algorithmic contributions. With the length of this paper in mind, I would suggest shortening Sections 6 and 7 in order to focus more on numerical comparisons of the new algorithm and less on stochastic reaction networks. Details of the multiscale material could be moved into supplementary material or even into another manuscript.}{}

\comment{Regularization. In regards to the regularization terms in equation (3.2), the authors claim that "experimentation has shown that the choice $\beta=1$ is sufficient for most problems." I would like to see more discussion on this choice. With a sufficiently large number of samples, decreasing $\beta$ should result in a more accurate transport map, so why does a smaller value of $\beta$ not yield a more efficient algorithm? With $\beta=1$, is the map somehow able to capture enough of the posterior structure to enable effective importance sampling?}{}

\comment{Proposal distribution. The proposed transport map ETAIS algorithm is tested using local random walk proposals for each ensemble member. However, for accurate maps, an independent Gaussian proposal or some form of mixture proposal could potentially generate nearly iid samples of the target distribution. Is there a reason for sticking with the random walk proposals? I think adding a comparison of proposal types, even if only on the Rosenbrock toy problem, would help readers understand the importance of each part in the transport map ETAIS algorithm: the proposal, map, resampler, and ensemble size.}

\comment{Ensemble Size. How is the ensemble size chosen? The authors mention that a smaller ensemble can be used when the transport map is accurate, but are there any guidelines on how to choose the ensemble size given the map?}{}

\comment{Closing Remarks. I found the discussion (Section 8) ended rather abruptly. When reading the manuscript, I felt like I walked off the end of the manuscript. I think a sentence or two that ties things together would provide a better closing statement.}

\comment{I would prefer numbering all equations, but leave that to the author's discretion.}{}

\comment{Subscript in Algorithm 1 on line 3. Should the proposal be centered at $\theta_j$ or $\theta_i$}{}

\comment{Second equation on page 4. Should the sample be labeled with a $\hat{ }$?}{}

\comment{When discussing monotonicity of the map, it might be worth pointing out the monotone parameterizations introduced in "An Introduction to Sampling via Measure Transport" by Marzouk et al.}{}

\comment{In Section 4, the notation for proposed points on the reference space (marked with an apostrophe) and proposed points on the target space (marked with a $\hat{ }$) are different. It might be easier to follow if these were consistent, but that is up to the authors.}{}

\comment{Equation (6.2) $Y_j$ is used in the text but $X$ is used in the equation. Is that a typo?}{}

\comment{Sentence before equation (6.3). "variable" should be plural "variables"}{}

\section*{Referee 2 comments}

\comment{I do think that having these realistic applications is a strength—but I think the paper would be much easier to digest if the pieces were better woven together at the outset. On that note, the introduction takes a rather long time to get to the point of combining transport-ETIAS with partially observed stochastic reaction networks. Perhaps the necessary link here is “sloppiness,” mentioned on page 3?}{}

\comment{Also, it should be made more clear what exactly is new within the stochastic reaction networks portion of the paper}{}

\comment{In the numerical examples, what exactly is the “relative L2 error,” or $\delta_{L^2}$? I don’t see where this is defined, and I don’t know it as a standard MCMC or importance sampling diagnostic. Is it the relative $L_2$ error in the estimate of some posterior moment? If so, which moment? And relative to what? Similarly, can you define $\delta_{ESS}$ as a function of the ESS?}

\comment{Do the error versus \# samples plots (as in Figure 5.3, Figure 7.3, Figure 7.8) correspond to a single runs of MCMC? Or are they the error averaged over many independent chains (in
1
which case I think plotting all the runs in the same figure would be more illustrative). These are Monte Carlo schemes, so I expect the error in any given running estimate to be noisy.}{}

\comment{By \# of samples, in the plots for MCMC error, do you simply mean the number of MCMC iterations? Similarly, what is meant by \# of samples in the ETIAS error plots? Using “samples” without context is confusing, as there are many intermediate populations of samples at hand.}{}

\comment{To complement these plots, I think that a very useful and widely used diagnostic, applicable both to MCMC and ETIAS and their transport-accelerated versions, would be ESS per posterior density evaluation, or simply ESS per unit of computational time.}{}

\comment{I think there is some confusion about invertibility in various places.
In the discussion following Definition 3.2, it is stated that the space of all invertible maps might not contain an exact coupling. For absolutely continuous distributions with strictly positive densities, which seems to correspond to the cases here, I do not believe that this statement is correct. There is always a bijective map from one distribution to the other (and vice versa); indeed, there is an infinity of such maps. If one only considers maps with triangular structure, then there is a unique (up to sets of measure zero) invertible transformation. This is more or less the Knothe–Rosenblatt rearrangement; see one of Villani’s books, or [Bogachev, Kolesnikov, Medvedev 2005] for a more formal treatment.
Similarly, the statement in Section 8.2 about the true map being “invariably itself not in- vertible” seemed odd to me. If by the true map one means the KR rearrangement, then it is invertible. Maybe I am confused about what is meant by “true map?”}{}

\comment{ In comparing the transport MH and the transport-ETIAS algorithms, I think it is somehow not surprising that in many problems the latter performs better. But I also have a hunch that the difference may be due to “global” versus local proposals. Transport MH, as I understand the implementation here, is restricted to using local random-walk proposals in the reference space. But some of the larger efficiency gains in [46] resulted from the use of a Metropolis independence proposal on the reference space, in particular draws from a standard Gaussian, i.e., the entire reference measure. These global moves yielded excellent mixing when the map was sufficiently accurate, and were combined with local moves using a delayed rejection scheme. ETIAS is not exactly the same, but it does have something in common with a global proposal. More broadly, this reflects the close link between independence Metropolis algorithms and importance sampling. Some comments (please agree or disagree) on this front would be illuminating.}{}

\comment{What part of Section 6 is newly developed in this paper? Is the derivation of the exact likelihood in Section 6.1 new? Or is the approximation in Section 6.2 new? This relates to my broader comment above. It would be good to have a clearer picture of where exactly this paper’s key contributions lie, and even of their relative weight.}{}

\comment{Here are several suggestions for content that could be removed entirely or shortened, along with a few areas that could be expanded:
  \begin{itemize}
    \item The use of log-transformations to map strictly positive variables to the real line is a standard trick in MCMC, and has also been employed in previous work on transport. One could simply mention this and move forward. Thus I don’t see the need of including any results without the log preconditioner in problems where it is warranted. Log- preconditioning is well recognized as a good thing to do.
\item A lot of Section 3 could be abridged by appropriate referencing of [46].
\item Section 4 is quite terse, but the“second option” (Algorithm 3)—both proposing and resampling on the reference space—seems to be the cleaner algorithm and, if I understand the authors, the more effective one. So perhaps the first option can be mentioned in passing, and all the results can focus on the second? Is there any good reason to use Algorithm 2 instead of Algorithm 3?
\item The description of the ETIAS algorithm in Section 2 is very nice, but here an illustrative figure might help. \end{itemize}}{}

\comment{End of Section 2, “learn local covariances across the whole of the domain”: I believe I under- stand what is meant, but this awkwardly phrased}{}

\end{document}

### 2019-11-01 Resubmission/TPAIS.tex
Resubmission folder and response.tex

\documentclass[final]{siamltex}
%test change
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks,soul}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{tikz,hhline}
\usepackage{tabularx}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

In this section, we give details on how we have adapted the method
proposed in \cite{parno2014transport} to couple the target,
$\mu_{\theta}$, with the reference distribution, $\mu_r$. In many
applications ***REFS*** inverse problems arise which lead to posterior
distributions from which it is challenging to sample efficiently. This
can be due to multimodality, strong correlations between parameters,
or vastly different scales of sensitivity in different directions, for
example the concept of sloppy models. Standard random walks become
inefficient in this context since the proposal distributions do not
take into account these types of challenging structures in the
density, leading to a large number of proposals being made in low
density regions.

One way that this can be tackled is to implement more complex MCMC
methods which are able to explore the state space more
efficiently. For example the Metropolis-Adjusted Langevin Algorithm (MALA) or Hamiltonian Monte
Carlo (HMC) methods use gradient information about the target density
in order to make proposals in better directions that RWMH. Riemannian
manifold MCMC methods can use additionally approximations of the
Hessian of the target in order to make very large moves with a good
probability of being accepted.

Another approach is the simplification of the posterior density
itself. In \cite{parno2014transport}, the authors proposed the
approximation of an invertible map which exactly couples the postieror distribution
$\mu_{\theta}$ to a reference measure $\mu_r$. Given such a map,
$T\in\mathcal{T}$, $T\colon \mathcal{X}\rightarrow\mathbb{R}^d$, we can then draw a proposal sample from $\pi_r$, before mapping these samples back onto target space using $T^{-1}$. These proposed samples are distributed according to the target distribution. On target space, the proposal distribution is induced by the pullback of the standard Gaussian through the map,
\begin{equation}\label{eq:pullback}
	\tilde{\pi}(\theta) = \pi_r(T(\theta))|J_T(\theta)|,
\end{equation}
which is equal to the target density when $T$ is exact. Clearly we
require $T$ to be monotonic and have continuous first derivatives,
which means that our space of feasible transference plans,
$\mathcal{T}^\uparrow$, may not include the exact map. Since it is
unlikely that $T\in\mathcal{T}^\uparrow$, we are motivated to
formulate an optimisation problem as discussed in
Section~\ref{sec:intro_optim}.  ****PAUL, what is
$\mathcal{T}^\uparrow$, what is in Section~\ref{sec:intro_optim}*****

We attempt to find a deterministic coupling of two continuous probability distributions, $(\mu_\theta, \mu_r)$, such that $\mu_r = T\mu_\theta$ is satisfied and also that $T \in \mathcal{T}^\uparrow$. There may be infinitely many such couplings, so we look to find the coupling which minimises the difference between the proposal density and the target distribution, i.e. the distance between $\pi(\theta)$ and $\tilde{\pi}(\theta)$. In this case, we measure this distance using the Kullback-Leibler divergence.

The map does not necessarily satisfy the monotonicity condition when we optimise the cost function,
\[
	C(T) = D_\text{KL}(\pi\|\tilde{\pi}),
\]
so it must be enforced in the design of each map component. To guarantee these properties we carefully choose the basis functions for each component, and we restrict the map to be lower triangular, i.e. $T \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. This lower triangular map has the form,
\[
	T(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
		T_n(\theta_1, \dots, \theta_n) \end{bmatrix},
\]
where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. We also assume that the target and reference probability densities are absolutely continuous on
$\mathbb{R}^d$ which guarantees the maps existence.

@@

With these constraints in mind we formulate the optimisation problem explicitly. The cost function
is chosen to be the Kullback-Leibler divergence between the posterior density and the pullback density,
\[
	D_\text{KL}(\pi\|\tilde{\pi}) =
		\mathbb{E}_\pi\left[\log\left(\frac{\pi(\theta)}{\tilde{\pi}(\theta)}\right)\right].
\]
Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map, the divergence becomes
\[
	D_\text{KL}(\pi\|\tilde{\pi}) = \mathbb{E}_\pi\left[\log\pi(\theta) - \log\pi_r(\tilde{T}(\theta)) -
		\log\left|J_{\tilde{T}}(\theta)\right|\right].
\]
We note the posterior density is constant in $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function. This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown, however this is exactly the scenario for which we would turn to MCMC methods for a solution.

To find the best coupling, $\tilde{T} \in \mathcal{T}^{\ltri}$, we solve the optimisation problem,
\[
	\tilde{T} = \arg\min_{T \in \mathcal{T}^{\smallltri}} \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
		\log\left|J_T(\theta)\right|\right]
\]
which has a unique solution. However we would like to ensure that this solution is regular and monotonic. For this
reason we regularise the problem further. The optimisation problem now takes the form
\begin{equation}\label{eq:gen_map_optim}
	\tilde{T} = \arg\min_{T\in\mathcal{T}^{\smallltri}} \left[
		 \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
		\log\left|J_T(\theta)\right|\right] + \beta\mathbb{E}(T(\theta)- \theta)^2 \right].
\end{equation}
This parameter $\beta$ does not need to be tuned, experimentation has shown that the choice
$\beta=1$ is sufficient for most problems. The form of the penalisation term promotes maps which are
close to the identity, and hence are likely to be monotonic.

@@

Before we continue with the derivation of the optimisation problem, we consider the structure
of the map in more detail. The lower lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density as well as the inverse of the map, $T^{-1}$. The Jacobian of $T$ is a lower triangular matrix,
\[
	DT(\theta) = \begin{bmatrix}
		\partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta)\\
		\vdots & \ddots & \vdots \\
		\partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta)
	\end{bmatrix} = \begin{bmatrix}
		\partial_{\theta_1} T_1(\theta) & \dots & 0\\
		\vdots & \ddots & \vdots \\
		\partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta)
	\end{bmatrix}
\]
since $\partial_{\theta_n} T_k(\theta) = 0$ for all $n > k$. This lower triangular structure means that the determinant of the Jacobian is a product of the diagonal elements which, when we take logs, becomes
\begin{equation}\label{eqn:separable_jacobian}
	\log\left|J_T(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} T_i(\theta).
\end{equation}
Here we note that this term is separable in terms of the dimension $i$.

Inverting $T$ is simplified by the lower triangular structure in the sense that each dimension, $i$, we need only invert a univariate polynomial. This simplification comes about because we invert each dimension of the map independently, beginning with $T_1$ which is a function of only $\theta_1$. Finding the root of $T_1(\theta_1) = r_1$, gives us the value of $\theta_1$ which we can then use to find the root of $T_2(\theta_2;\theta_1) = r_2$, and so on. These roots, $r_1,\dots,r_d$, are restricted to be real since $\mathcal{X} \subset \mathbb{R}^d$, which allows us to find the unique point, $\theta$, which corresponds to the point $r$.

We require that the first derivatives of the map are continuous, which is easy to enforce by the choice of basis functions. Here we assume that the map will be built from a family of orthogonal polynomials, $\mathcal{P}(\theta)$, although not necessarily orthogonal with respect to the target distribution. Each component of the map is defined as a multivariate polynomial expansion,
\begin{equation}\label{eq:map_defn}
	\tilde{T}_i(\theta; \gamma_i) = \sum\limits_{\mathbf{j}\in\mathcal{J}_i} \!
\gamma_{i,\mathbf{j}}\psi_\mathbf{j}(\theta).
\end{equation}
The parameter $\gamma_i$ is a vector of coefficients in $\mathbb{R}^{M_i}$. Each component of $\gamma_i$ corresponds to a basis function
$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in \mathbb{N}_0^d$, here $d$ is the dimension of
$\theta$. These multi-indices are elements of the multi-index set $\mathcal{J}_i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
\[
	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
\]
and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$. Since $\tilde{T}$ is lower triangular, a multi-index $\mathbf{j}\in\mathcal{J}_i$ only contains entries for univariate polynomials in $\theta_k$ for $k\leq i$.

The cardinalities of the multi-index sets $M_i = \text{card}(\mathcal{J}_i)$ give the number of unknowns in our
optimisation problem, and so we would like to keep this number as small as possible. One option is
to use polynomials of total order $p$,
\[
	\mathcal{J}_i^\text{TO} = \left\{\mathbf{j}:\|\mathbf{j}\|_1 \leq p, j_k = 0\ \forall k > i\right\},
\]
which is optimal in terms of the amount of information captured by the map about the target. The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = {i+p \choose p}$ which increases rapidly in $d$ and $p$. Smaller optimisation problems can be produced by constructing subsets of $\mathcal{J}_i^\text{TO}$. These index sets are discussed further
in~\cite{parno2014transport}, here we stick with polynomials of total order $p$ since we work with low dimensional problems with the PAIS algorithm.

>>>>

\documentclass[final]{siamltex}
%test change
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks,soul}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{tikz,hhline}
\usetikzlibrary{positioning,automata}
\usetikzlibrary{arrows,fit}
\usetikzlibrary{decorations.pathmorphing}
\usepackage{tabularx}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\let\oldnl\nl% Store \nl in \oldnl
\newcommand{\nonl}{\renewcommand{\nl}{\let\nl\oldnl}}% Remove line number for one line

@@

So far we have seen that the PAIS algorithm is efficient for low dimensional posterior sampling.
PAIS is particularly useful if the posterior is multimodal, or inconveniently shaped such as the
Rosenbrock density discussed in Chapter~\ref{sec:LPAIS}. However, we have had trouble when
attempting to increase the dimension of the parameter space. This was seen when we attempted to
recover the initial condition for the wave equation in Chapter~\ref{sec:PAIS}.

In this Chapter we look at a method which will allow us to improve efficiency in our proposal distributions. The ideal sampling
situation in MCMC is to explore an $d$-dimensional Gaussian distribution (or other standard distribution) using a near proportional
proposal distribution. We therefore aim to transform an arbitrary distribution, $\mu_\theta$, in to a Gaussian distribution, which we can then sample from efficiently, and finally transform this sample back into a sample from $\mu_\theta$. In this way we hope to be able to sample from any distribution $\mu_\theta$ with a similar efficiency as we obtain when sampling from a Gaussian distribution. In particular for the PAIS algorithm, resampling schemes are more efficient in higher
dimensions if the dimensions are uncorrelated. The work of Moselhy and
Marzouk~\cite{el2012bayesian}, Parno and Marzouk~\cite{parno2014transport} and
Parno~\cite{parno2015transport} provides a method of constructing such a map as part of an adaptive
Metropolis-Hastings algorithm.

%Given some constraints it is computationally tractable to adaptively construct invertible maps
%during an MCMC simulation. These maps construct a continuous coupling between a target and
%reference distribution, where the latter is chosen to be a standardised Gaussian. How close this reference
%distribution is to Gaussian depends on three considerations; firstly the constraints applied. These
%constraints, discussed later, ensure that the map exists, is unique, and is invertible. The
%necessity for the map to be invertible, and preferrably easily invertible, restricts the space of
%possible couplings to the point where it is unlikely that the exact map is included in our space of
%feasible transference plans. Secondly, for our purposes, the map takes the form of a polynomial
%chaos expansion which requires infinitely many terms to exactly represent the coupling. This is of course not possible on current hardware. Finally we
%will be constructing this map using an optimisation problem where the objective function depends on
%the sample path of our MCMC algorithm. This sample does not represent the target exactly until $N\rightarrow\infty$, and so the expectation is only calculated approximately.
%
%While the restrictions on the map are essential for well-posedness, the approximate map which
%results from these constraints can be made arbitrarily accurate by increasing the number of basis
%functions used in the expansion, and increasing the number of samples used in the optimisation
%problem. In practice it appears that we need not aim to achieve a high accuracy solution. Excellent
%results have been observed using monomial basis functions of total order 3, and we need a sample of
%size of only $\mathcal{O}(10^3)$ to obtain a well converged map.

In~\cite{el2012bayesian} the transport map was introduced to provide a transformation from the prior
distribution to the posterior distribution, the idea being that one could draw a moderately sized
sample from the prior distribution and use this sample to approximate a map onto the target space.
Once this map was known to the desired accuracy a larger sample from the prior could be used to
investigate the posterior distribution. This
methodology was adapted in~\cite{parno2014transport} to form a new proposal method for MH
algorithms. In this case, rather than transforming a sample from the prior into a sample from the target
distribution, the map transforms a sample from the posterior onto a reference space.
The reference density is chosen to allow efficient proposals using a simple proposal
distribution such as a Gaussian centred at the previous state. Proposed states can then be mapped back into a sample from the posterior by applying the inverse of the transport map.

Proposing new states in this way allows us to make large steps around complex probability distributions.
It is also feasible in this framework to assume that the reference density is close enough to a standard Gaussian that we can efficiently propose moves using a proposal distribution which is independent of the current state, e.g. choose $q(\theta) = \mathcal{N}(0,I_n)$.

In this chapter we introduce the derivation of the map discussed, show how it can be adapted to work
with a weighted sample, and that mixture distributions as used in the PAIS algorithm can more
efficiently approximate the reference space and so lead to higher overall effective sample sizes.
Further we show that the algorithm arising from the combination of these methods can lead to more
efficient sampling than either can achieve on its own.

\section{The Transport Map}

In this section we outline the methodology in \cite{parno2014transport} for coupling the target,
$\mu_{\theta}$, with the reference distribution, $\mu_r$.
We also show how the map can be constructed using a weighted sample and hence how we can incorporate the map into the PAIS algorithm.

\begin{dfn}[Transport Map $T$]
	Given a probability distribution $\mu_\theta$ with support $\mathcal{X}$.
	A transport map is a function $T\colon \mathcal{X} \rightarrow \mathbb{R}^d$ such that
	\[
		\mu_\theta(A) = \mu_r(T(A)) \quad \text{for any } A \subset \mathcal{X}.
	\]
\end{dfn}

\begin{dfn}[Exact Transport Map $T$]
	A transport map $T$ is exact if the {\it pullback} of the standard Gaussian density, $\phi(\cdot)$,
	\begin{equation}\label{eq:pullback}
		\tilde{\pi}(\theta) = \phi(T(\theta))|J_T(\theta)|,
	\end{equation}
	is equal to the target density $\pi(\theta)$ for all $\theta \in \mathcal{X}$. The pullback is defined in terms of the determinant of the Jacobian of $T$,
	\[
		|J_T(\theta)| = \text{det}\begin{bmatrix} \partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta) \\ \vdots & \ddots & \vdots \\ \partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta) \end{bmatrix}.
	\]
\end{dfn}

In the case that we have the exact map, $T\in\mathcal{T}$ where $\mathcal{T}$ is the space of all invertible maps, we are able to draw a sample from $\pi_r = \phi$, the density of the standardised Gaussian distribution, and map these samples back onto target space using $T^{-1}$. These proposed samples are distributed according to the target distribution.

\begin{dfn}[Target and Reference Space]
	The transport map pushes a particle from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target measure $\mu_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference measure $\mu_r$.
\end{dfn}

On target space, the proposal density is induced by the pullback of $\phi$ through $T^{-1}$.
Clearly the pullback only exists when $T$ is monotonic and has continuous first derivatives.
Not all maps satisfy these conditions, so we define a smaller space of maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$ which contains all feasible maps.
The exact map $T$ is not necessarily in $\mathcal{T}^\uparrow$, so we are motivated to formulate an optimisation problem as discussed in Section~\ref{sec:intro_optim}.

We attempt to find a deterministic coupling of two continuous probability distributions, $(\mu_\theta, \mu_r)$, such that $\mu_r = T\mu_\theta$ is satisfied and also that $T \in \mathcal{T}^\uparrow$. There may be infinitely many such couplings, so we look to find the coupling which minimises the KL-divergence between the density of $\mu_{\theta}$ and the pullback of the density of $\mu_{r}$, i.e. the distance between $\pi(\theta)$ and $\tilde{\pi}(\theta)$.

When we optimise the cost function
\[
	C(T) = D_\text{KL}(\pi\|\tilde{\pi}),
\]
numerically, it is not guaranteed that the resulting map will be invertible, so it must be enforced in the design of each map component. To ensure invertibility we restrict the map to be lower triangular, i.e. $\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. This lower triangular map has the form,
\[
	T(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
		T_n(\theta_1, \dots, \theta_n) \end{bmatrix},
\]
where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. We assume that the target and reference probability densities are absolutely continuous on
$\mathbb{R}^d$. Under this formulation we are guaranteed a unique invertible map $\tilde{T}$ with the property that
\[
	\tilde{T}\mu_{\theta} \approx \mu_r.
\]
Relaxing the equality constraint to finding the approximate map $\tilde{T} \in \mathcal{T}^{\ltri}$ which optimises $C(T)$, allows us to formulate an efficient sampling method, in terms of the linear algebra required, and experimentation has shown that the pushforward of the target is close enough to Gaussian to achieve efficient sampling.

@@

With these constraints in mind we formulate the optimisation problem explicitly. The cost function
is chosen to be the Kullback-Leibler divergence between the posterior density and the pullback density,
\[
	D_\text{KL}(\pi\|\tilde{\pi}) =
		\mathbb{E}_\pi\left[\log\left(\frac{\pi(\theta)}{\tilde{\pi}(\theta)}\right)\right].
\]
This divergence results in some nice properties which we will explore in the following derivation. The KL divergence is not a true metric since it is not symmetric, however it is commonly used to measure the distance between probability distributions due to it's relatively simple form, and because it provides a bound for the square of the Hellinger distance by Pinsker's inequality~\cite{pinsker1960information},
\[
	D_{KL}(p\|q) \geq D_H^2(p,q),
\]
which is a true metric between probability distributions $p$ and $q$.
Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map $\tilde{T}$, the divergence becomes
\[
	D_\text{KL}(\pi\|\tilde{\pi}) = \mathbb{E}_\pi\left[\log\pi(\theta) - \log\pi_r(\tilde{T}(\theta)) -
		\log\left|J_{\tilde{T}}(\theta)\right|\right].
\]
We note the posterior density is independent of $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function. This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown. However this is exactly the scenario for which we would turn to MCMC methods for a solution.

To find the best coupling, $\tilde{T} \in \mathcal{T}^{\ltri}$, we solve the optimisation problem,
\[
	\tilde{T} = \arg\min_{T \in \mathcal{T}^{\smallltri}} \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
		\log\left|J_T(\theta)\right|\right]
\]
which has a unique solution since the cost function is convex.

We also include a regularisation term which is required for reasons which will become clear later. The optimisation problem now takes the form
\begin{equation}\label{eq:gen_map_optim}
	\tilde{T} = \arg\min_{T\in\mathcal{T}^{\smallltri}} \left[
		 \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
		\log\left|J_T(\theta)\right|\right] + \beta\mathbb{E}(T(\theta)- \theta)^2 \right].
\end{equation}
This parameter $\beta$ does not need to be tuned, experimentation has shown that the choice
$\beta=1$ is sufficient for most problems. The form of the penalisation term promotes maps which are
closer to the identity.

@@

Before we continue with the derivation of the optimisation problem, we consider the structure
of the map in more detail. The lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density, as well as the inverse of the map, $\tilde{T}^{-1}$. The Jacobian of $\tilde{T}$ is a lower triangular matrix,
\[
	J_T(\theta) = \begin{bmatrix}
		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & \partial_{\theta_d} \tilde{T}_1(\theta)\\
		\vdots & \ddots & \vdots \\
		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
	\end{bmatrix} = \begin{bmatrix}
		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & 0\\
		\vdots & \ddots & \vdots \\
		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
	\end{bmatrix}
\]
since $\partial_{\theta_n} \tilde{T}_k(\theta) = 0$ for all $n > k$. This lower triangular structure means that the determinant of the Jacobian is a product of the diagonal elements which, when we take logs, becomes
\begin{equation}\label{eqn:separable_jacobian}
	\log\left|J_{\tilde{T}}(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} \tilde{T}_i(\theta).
\end{equation}
Here we note that this term is separable in terms of the dimension $i$.

%Inverting $T$ is simplified by the lower triangular structure in the sense that in each dimension, $i$, we need only invert a univariate polynomial. This simplification comes about because we invert each dimension of the map independently, beginning with $T_1$ which is a function of only $\theta_1$. Finding the root of $T_1(\theta_1) = r_1$, gives us the value of $\theta_1$ which we can then use to find the root of $T_2(\theta_2;\theta_1) = r_2$, and so on. These roots, $r_1,\dots,r_d$, are restricted to be real since $\mathcal{X} \subset \mathbb{R}^d$, which allows us to find the unique point, $\theta$, which corresponds to the point $r$.

Inverting $\tilde{T}$ at a point $r$ is simplified by the lower triangular structure of the map. The map component $\tilde{T}_1(\theta)$ is a univariate polynomial in $\theta_1$, so we can find the inverse of this function by solving the equation $T_1(\theta_1) = r_1$. This inversion tells us the value of $\theta_1$, which means the next component is again a univariate polynomial, $T_2(\theta_2; \theta_1)=r_2$. We can then perform $d$ root finding problems instead of a full $d$ dimensional non-linear solve.

We require that the first derivatives of the map are continuous, which is easy to enforce by the choice of basis functions. Here we assume that the map will be built from a family of orthogonal polynomials, $\mathcal{P}(\theta)$, not necessarily orthogonal with respect to the target distribution. Each component of the map is defined as a multivariate polynomial expansion,
\begin{equation}\label{eq:map_defn}
	\tilde{T}_i(\theta; \gamma_i) = \sum\limits_{\mathbf{j}\in\mathcal{J}_i} \!
\gamma_{i,\mathbf{j}}\psi_\mathbf{j}(\theta).
\end{equation}
The parameter $\gamma_i$ is a vector of coefficients in $\mathbb{R}^{M_i}$. Each component of $\gamma_i$ corresponds to a basis function
$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in \mathbb{N}_0^d$. These multi-indices are elements of the multi-index set $\mathcal{J}_i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
\[
	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
\]
and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$. Since $\tilde{T}$ is lower triangular, a multi-index $\mathbf{j}\in\mathcal{J}_i$ only contains entries for univariate polynomials in $\theta_k$ for $k\leq i$.
